On the probability of finding a nonphysical solution through shadowing

### 2020-10-26 main.tex
Fix type errors up to page 8

\institute{Q. Wang \at 
        Department of Aeronautics and Astronautics, Center for Computational Science and Engineering, MIT, Cambridge, MA 02139, USA.
              \email{qiqi@mit.edu}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
           N. Chandramoorthy \at Department of Mechanical Engineering, Center for Computational Science and Engineering, MIT, Cambridge, MA 02139, USA. \email{nishac@mit.edu}
}

@@

In this section, we first describe mathematically what distinguishes a physical
and a nonphysical solution.  We also explain theoretically why it is unlikely
to observe a nonphysical solution in an experiment.  We then give a few
examples of these almost-never-observed nonphysical solutions.

\subsection{Physical solutions}
What we typically call ``physical'' solutions to a governing equation satisfy the following
two criteria: 
\begin{enumerate}
    \item Long-time-averaged, ``statistical'' quantities converges as time goes to infinity.
    Consider $u(t)$ as the solution to a chaotic governing equation, then for a regular observable
    of interest $J(u(t))$,
    \begin{equation} \label{longtimeavg}
        \lim_{T\to\infty}\frac1T \int_0^T J(u(t))\,dt
    \end{equation}
    exists.
    \item For almost any small perturbation to the initial condition $u(0)\to u'(0)$, the perturbed
    solution $u'(t)$ should have the same statistics, i.e.,
    \begin{equation}
        \lim_{T\to\infty}\frac1T \int_0^T J(u(t))\,dt
      = \lim_{T\to\infty}\frac1T \int_0^T J(u'(t))\,dt
    \end{equation}
\end{enumerate}

\begin{figure} \centering
\includegraphics[width=0.48\textwidth]{lorenz_trajectory_1000.0.png}
\hspace{0.02\textwidth}
\includegraphics[width=0.48\textwidth]{lorenz_ensemble_10.png}
\caption{
Left: An ensemble of initial conditions distributed uniformly in the box $u_1\in[0,1], u_2\in[0,1], u_3\in[28,39]$ 
after 10 time units of evolution, shown in in the $u_1-u_3$ plane.
Right: distribution of a trajectory of $1000$ time units in length.
In both plots, the color represent the number of samples in a
$2048\times2048$ uniform grid.  The trajectory is sampled every 0.001 time units.
}
\label{fig:lorenz_ergodicity1}
\end{figure}
\begin{figure} \centering
\includegraphics[width=0.48\textwidth]{lorenz_trajectory_10000.0.png}
\hspace{0.02\textwidth}
\includegraphics[width=0.48\textwidth]{lorenz_ensemble_15.png}
\caption{
Left: distribution of the same ensemble as in Figure \ref{fig:lorenz_ergodicity1}
after another 5 time units of evolution.
Right: distribution of the same trajectory as in Figure \ref{fig:lorenz_ergodicity1}
evolved for $10,000$ time units.
}
\label{fig:lorenz_ergodicity2}
\end{figure}
\begin{figure} \centering
\includegraphics[width=0.48\textwidth]{lorenz_trajectory_1000000.0.png}
\hspace{0.02\textwidth}
\includegraphics[width=0.48\textwidth]{lorenz_ensemble_50.png}
\caption{
Left: distribution of the same ensemble as in Figures \ref{fig:lorenz_ergodicity1}
and \ref{fig:lorenz_ergodicity2} after a total of 50 time units of evolution.
Right: distribution of the same trajectory as in Figures \ref{fig:lorenz_ergodicity1}
and \ref{fig:lorenz_ergodicity2} evolved for $100,000$ time units.
}
\label{fig:lorenz_ergodicity3}
\end{figure}

The left column of Figures \ref{fig:lorenz_ergodicity1}-\ref{fig:lorenz_ergodicity3}
illustrates a physical solution of the Lorenz system,
\begin{equation}
\label{eqn:lorenz}
\frac{du}{dt} =
    \frac{d}{dt}\begin{pmatrix}u_1\\u_2\\u_3\end{pmatrix} =
    \begin{pmatrix}
        \sigma (u_2-u_1) \\
        u_1(\rho -u_3) - u_2 \\
        u_1\,u_2 - \beta\,u_3
\end{pmatrix}
\end{equation}
where $\sigma=10,\rho=28$, and $\beta=8/3$.  The solution starts at the initial
condition $u(0)=(0.01, 0.01, 28)$.  After a long time evolution, we observe that
the solution has visited a large portion of the $u_1-u_3$ plane with a varying, but well-defined, frequency.  We can use a probability distribution, $\mu$, to quantify
how frequently a physical solution visits portions the phase space.
Specifically, for a subset $A$ of the phase space, $\mu(A)$ measures the
fraction of time a very long physical solution spends inside the subset $A$.
With this distribution defined, the infinite-time average of any quantity $J(u)$ can be represented as an average of $J$ over the entire phase space $U$, weighted
by this statistical distribution $\mu$.  Mathematically,
\begin{equation}
    \frac1T \int_0^T J(u(t))\,dt \xrightarrow{T\to\infty}
    \int_U J(u)\: d\mu(u)
\end{equation}

Remarkably, the distribution $\mu$ not only characterizes the history of
a long physical solution, but also describes the settled state of
an ensemble of solutions.
This is illustrated on the right column of Figures \ref{fig:lorenz_ergodicity1}--\ref{fig:lorenz_ergodicity3}.
We generate these plots by starting from an ensemble of about one billion initial
conditions, randomly and uniformly spaced in a small three-dimensional box.  All these billion solutions are evolved by solving Eq. \ref{eqn:lorenz} for 10, 15, and 50 time units, to obtain the plots. 
We observe that, as time evolves, the ensemble spreads over an increasingly
larger portion of the $u_1-u_3$ plane. After a long time, the ensemble settles into a set known as an \emph{attractor} that does not change with time, and contains the \emph{unstable manifold}, which appears as filaments forming the attractor. The distribution of the ensemble on the attractor becomes identical to the distribution of a single, very long physical
solution, which is also contained in the attractor (Figure \ref{fig:lorenz_ergodicity3}).

@@

Because periodic solutions have a more limited extent in the phase space, their
statistics are different from physical solutions.
Here we illustrate the difference using the mean of two
quantities of interest
\begin{align}
    J_1(u) = u_3\;,\quad J_2(u) = e^{-\dfrac{u_3^2}2}.
\end{align}
Table \ref{tab:lorenz_periodic_stats} shows these quantities
of interest averaged over the three periodic solutions shown in Figure
\ref{fig:lorenz_periodic}, compared against those averaged over a
physical solution.  Here, Periodic \#1, \#2, and \#3 correspond to
the solid, dashed, and dotted lines, respectively.
We observe from the table that although the mean of $J_1$ over the periodic
solutions are different but comparable to the mean over physical solutions.  The mean of $J_2$ over the periodic solutions, on the other hand,
is orders of magnitude different from that of the physical solutions.
These differences disqualify the periodic solutions from being physical.
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c|c}
             & Periodic \#1 & Periodic \#2
             & Periodic \#3 & Physical solutions \\
    \hline
        $J_1$& 23.05 &
               23.19 &
               23.37 &
               23.67 \\
        $J_2$&$4\times 10^{-35}$&
              $7\times 10^{-28}$&
              $3\times 10^{-22}$&
              $1.58\times 10^{-05}$
    \end{tabular}
    \caption{Comparison of statistics computed from periodic solutions of the Lorenz equation with the statistics computed from physical solutions}
    \label{tab:lorenz_periodic_stats}
\end{table}

@@

\begin{figure}\centering
\includegraphics[width=0.48\textwidth]{lorenz_zmax_10_28_2.6666666666666665.png}
\hspace{0.02\textwidth}
\includegraphics[width=0.48\textwidth]{lorenz_zmax2_10_28_2.6666666666666665.png}
\caption{L: the Lorenz map.  The x-axis is the $n$-th local maximum of $u_3(t)$ over a
long solution; the y-axis is the $(n+1)$-th local maximum of $u_3(t)$.  The intersection of this curve
		with the dashed line ($y=x$) indicates the initial condition for the solid line
in Figure \ref{fig:lorenz_periodic}.
R: the Lorenz map iterated twice.  The x-axis is the $n$-th local maximum
of $u_3(t)$; the y-axis is the $(n+2)$-th maximum.  The intersections with the diagonal dashed
line indicate the initial conditions for both the dotted and dashed lines in
Figure \ref{fig:lorenz_periodic}.}
\label{fig:lorenz_map}
\end{figure}
\nisha{In Figure 4, the solid line appears only to have one local maximum, which is shown as the fixed point on the left plot of this figure. The two fixed points of the 2-time-iterate of the Lorenz map, shown on the right plot of this figure, both appear to correspond to the local maxima of the dashed curve in Figure 4.}
The Lorenz map provides us a tool to find as many periodic solutions as we want.
By intersecting the map with a diagonal line, we can find a local maximum of $u_3(t)$
for which the next maximum is almost the same value.  We then look up the values of
$u_1(t)$ and $u_2(t)$ when $u_3(t)$ achieved this maximum.  This gives us an initial
condition starting from which the solution is nearly periodic.   We can similarly
intersect the second iterate of the Lorenz map (right plot of Figure \ref{fig:lorenz_map},)
and the third iterate, etc, with a diagonal line to find increasingly
complex periodic solutions.

>>>>

\institute{           %  \\
%             \emph{Present address:} of F. Author  %  if needed
           Center for Computational Science and Engineering, MIT, Cambridge, MA 02139, USA. \email{nishac@mit.edu 
           }
}

@@

In this section, we first describe what distinguishes, mathematically, a physical
and a nonphysical solution.  We also explain why it is theoretically unlikely
to observe a nonphysical solution in an experiment.  We then give a few
examples of these almost-never-observed nonphysical solutions.

\subsection{Physical solutions}
What we typically call ``physical'' solutions to a governing equation satisfy the following
two criteria: 
\begin{enumerate}
    \item Time-averaged quantities converge in the limit of infinite averaging time.
    Consider $u(t)$ as the solution to a chaotic governing equation, then for a regular observable
    of interest $J(u(t))$,
    \begin{equation} \label{longtimeavg}
        \lim_{T\to\infty}\frac1T \int_0^T J(u(t))\,dt
    \end{equation}
    exists.
    \item For almost any small perturbation to the initial condition $u(0)\to u'(0)$, the perturbed
    solution $u'(t)$ should have the same statistics, i.e.,
    \begin{equation}
        \lim_{T\to\infty}\frac1T \int_0^T J(u(t))\,dt
      = \lim_{T\to\infty}\frac1T \int_0^T J(u'(t))\,dt
    \end{equation}
\end{enumerate}

\begin{figure} \centering
\includegraphics[width=0.48\textwidth]{lorenz_trajectory_1000.0.png}
\hspace{0.02\textwidth}
\includegraphics[width=0.48\textwidth]{lorenz_ensemble_10.png}
\caption{
L: An ensemble of initial conditions distributed uniformly in the box $u_1\in[0,1], u_2\in[0,1], u_3\in[28,39]$ 
after 10 time units of evolution, shown on the $u_1-u_3$ plane.
R: distribution of a trajectory of $1000$ time units in length.
In both plots, the color represent the number of samples in a
$2048\times2048$ uniform grid.  The trajectory is sampled every 0.001 time units.
}
\label{fig:lorenz_ergodicity1}
\end{figure}
\begin{figure} \centering
\includegraphics[width=0.48\textwidth]{lorenz_trajectory_10000.0.png}
\hspace{0.02\textwidth}
\includegraphics[width=0.48\textwidth]{lorenz_ensemble_15.png}
\caption{
L: distribution of the same ensemble as in Figure \ref{fig:lorenz_ergodicity1}
after another 5 time units of evolution.
R: distribution of the same trajectory as in Figure \ref{fig:lorenz_ergodicity1}
evolved for $10,000$ time units.
}
\label{fig:lorenz_ergodicity2}
\end{figure}
\begin{figure} \centering
\includegraphics[width=0.48\textwidth]{lorenz_trajectory_1000000.0.png}
\hspace{0.02\textwidth}
\includegraphics[width=0.48\textwidth]{lorenz_ensemble_50.png}
\caption{
L: distribution of the same ensemble as in Figures \ref{fig:lorenz_ergodicity1}
and \ref{fig:lorenz_ergodicity2} after a total of 50 time units of evolution.
R: distribution of the same trajectory as in Figures \ref{fig:lorenz_ergodicity1}
and \ref{fig:lorenz_ergodicity2} evolved for $100,000$ time units.
}
\label{fig:lorenz_ergodicity3}
\end{figure}

The left column of Figures \ref{fig:lorenz_ergodicity1}-\ref{fig:lorenz_ergodicity3}
illustrates a physical solution of the Lorenz system,
\begin{equation}
\label{eqn:lorenz}
\frac{du}{dt} =
    \frac{d}{dt}\begin{pmatrix}u_1\\u_2\\u_3\end{pmatrix} =
    \begin{pmatrix}
        \sigma (u_2-u_1) \\
        u_1(\rho -u_3) - u_2 \\
        u_1\,u_2 - \beta\,u_3
\end{pmatrix}
\end{equation}
where $\sigma=10,\rho=28$, and $\beta=8/3$.  The solution starts at the initial
condition $u(0)=(0.01, 0.01, 28)$.  After a long time evolution, we observe that
the solution has visited a large portion of the $u_1-u_3$ plane with a varying, but well-defined, frequency.  We can use a probability distribution, $\mu$, to quantify
how frequently a physical solution visits portions of the phase space.
Specifically, for a subset $A$ of the phase space, $\mu(A)$ measures the
fraction of time a very long physical solution spends inside the subset $A$.
With this distribution defined, the infinite-time average of any quantity $J(u)$ can be represented as an average of $J$ over the entire phase space $U$, weighted
by this statistical distribution $\mu$.  Mathematically,
\begin{equation}
    \frac1T \int_0^T J(u(t))\,dt \xrightarrow{T\to\infty}
    \int_U J(u)\: d\mu(u)
\end{equation}

Remarkably, the distribution $\mu$ not only characterizes the history of
a long physical solution, but also describes the settled state of
an ensemble of solutions.
This is illustrated on the right column of Figures \ref{fig:lorenz_ergodicity1}--\ref{fig:lorenz_ergodicity3}.
We generate these plots by starting from an ensemble of about one billion initial
conditions, randomly and uniformly spaced in a small three-dimensional box.  All these billion solutions are evolved by solving Eq. \ref{eqn:lorenz} for 10, 15, and 50 time units, to obtain the plots. 
We observe that, as time evolves, the ensemble spreads over an increasingly
larger portion of the $u_1-u_3$ plane. After a long time, the ensemble settles into a time-invariant \emph{attractor} that contains the \emph{unstable manifold}, which appears as filaments forming the attractor. The probability distribution of the ensemble on the attractor becomes identical to the distribution of a single, very long physical
solution, which is also contained in the attractor (Figure \ref{fig:lorenz_ergodicity3}).

@@

Because periodic solutions have a more limited extent in the phase space, their
statistics are different from physical solutions.
Here we illustrate the difference using the mean of two
quantities of interest
\begin{align}
    J_1(u) = u_3\;,\quad J_2(u) = e^{-\dfrac{u_3^2}2}.
\end{align}
Table \ref{tab:lorenz_periodic_stats} shows these quantities
of interest averaged over the three periodic solutions shown in Figure
\ref{fig:lorenz_periodic}, compared against those averaged over a
physical solution.  Here, Periodic \#1, \#2, and \#3 correspond to
the solid, dashed, and dotted lines, respectively.
We observe from the table that the mean of $J_1$ over the periodic
solutions are different but comparable to the mean over physical solutions.  The mean of $J_2$ over the periodic solutions, on the other hand,
is orders of magnitude different from that of the physical solutions.
These differences disqualify the periodic solutions from being physical.
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c|c}
             & Periodic \#1 & Periodic \#2
             & Periodic \#3 & Physical solutions \\
    \hline
        $J_1$& 23.05 &
               23.19 &
               23.37 &
               23.67 \\
        $J_2$&$4\times 10^{-35}$&
              $7\times 10^{-28}$&
              $3\times 10^{-22}$&
              $1.58\times 10^{-05}$
    \end{tabular}
    \caption{Comparison of statistics computed from periodic solutions of the Lorenz equation with the statistics computed from physical solutions}
    \label{tab:lorenz_periodic_stats}
\end{table}

@@

\begin{figure}\centering
\includegraphics[width=0.48\textwidth]{lorenz_zmax_10_28_2.6666666666666665.png}
\hspace{0.02\textwidth}
\includegraphics[width=0.48\textwidth]{lorenz_zmax2_10_28_2.6666666666666665.png}
\caption{L: the Lorenz map.  The x-axis is the $n$-th local maximum of $u_3(t)$ over a
long solution; the y-axis is the $(n+1)$-th local maximum of $u_3(t)$.  The intersection of this curve
		with the dashed line ($y=x$) indicates the initial condition for the solid line
in Figure \ref{fig:lorenz_periodic}.
R: the Lorenz map iterated twice.  The x-axis is the $n$-th local maximum
of $u_3(t)$; the y-axis is the $(n+2)$-th maximum.  The intersections with the diagonal dashed
line indicate the initial conditions for both the dotted and dashed lines in
Figure \ref{fig:lorenz_periodic}.}
\label{fig:lorenz_map}
\end{figure}
The Lorenz map provides us a tool to find as many periodic solutions as we want.
By intersecting the map with a diagonal line, we can find a local maximum of $u_3(t)$
for which the next maximum is almost the same value.  We then look up the values of
$u_1(t)$ and $u_2(t)$ when $u_3(t)$ achieved this maximum.  This gives us an initial
condition starting from which the solution is nearly periodic.   We can similarly
intersect the second iterate of the Lorenz map (right plot of Figure \ref{fig:lorenz_map},)
and the third iterate, etc, with a diagonal line to find increasingly
complex periodic solutions.
